---
title: "Association Rule Mining"
author: "Katrina Arbogast"
output: html_document
---


```{r}
library(viridis)
library(arules)
library(TSP)
library(data.table)
#library(ggplot2)
#library(Matrix)
library(tcltk)
library(dplyr)
library(devtools)
library(purrr)
library(tidyr)
## DO THIS ONCE
## FIRST - you MUST register and log into github
## install_github("mhahsler/arulesViz")
## RE: https://github.com/mhahsler/arulesViz

##############
## IF YOUR CODE BREAKS - TRY THIS
##
## Error in length(obj) : Method length not implemented for class rules 
## DO THIS: 
## (1) detach("package:arulesViz", unload=TRUE)
## (2) detach("package:arules", unload=TRUE)
## (3) library(arules)
## (4) library(arulesViz)
###################################################################

## To see if you have tcltk run this on the console...
# capabilities()["tcltk"]
library(arulesViz)
```

#### Setting folder path and filnames

```{r}
#------------------------------------------------------------------------------
# Setting the folder path to the cleaned but not formatted data
or_weather_wildfires_comments_basket_path <- "./CleanData/or_weather_wildfires_cause_comments_basket.csv"
or_weather_wildfires_specific_basket_path <- "./CleanData/or_weather_wildfires_specific_cause_basket.csv"
news_headlines_basket_path <- "./CleanData/NewsHeadlines_basket.csv"

# Setting the filename
or_weather_wildfires_comments_filename <- "or_weather_wildfires_cause_comments"
or_weather_wildfires_specific_filename <- "or_weather_wildfires_specific_cause"
news_headlines_filename <- "NewsHeadlines"
#------------------------------------------------------------------------------
```

## Ingesting Cleaned Fire and Drought Data

```{r}
or_cause_comments <- read.csv(or_weather_wildfires_comments_basket_path)
or_specific_cause <- read.csv(or_weather_wildfires_specific_basket_path)
news_headlines <- read.csv(news_headlines_basket_path)
```


### Basket Data

```{r}
head(or_cause_comments)
head(or_specific_cause)
head(news_headlines)

# or_cause_comments
# or_specific_cause
# news_headlines
```

For this section of the project I have decided to leave the label columns, this is because I would like to analyze the relationship between them and the words that were found with them, as well as the associations between the words themselves. I think this could give a little insight into what kind of words are associated with fires in the news and what words are associated with the different causes of fires. 



## Getting the Transaction Data


```{r}
or_cause_comments_transactions <-
  read.transactions(or_weather_wildfires_comments_basket_path,
                    rm.duplicates = FALSE, 
                    format = "basket",  
                    sep=",",  
                    cols=NULL)

or_specific_cause_transactions <-
  read.transactions(or_weather_wildfires_specific_basket_path,
                    rm.duplicates = FALSE, 
                    format = "basket",  
                    sep=",",  
                    cols=NULL)

news_headlines_transactions <-
  read.transactions(news_headlines_basket_path,
                    rm.duplicates = FALSE, 
                    format = "basket",  
                    sep=",",  
                    cols=NULL)
```


## Apriori

```{r}
or_cause_comments_rules = arules::apriori(or_cause_comments_transactions,
                                          parameter = list(support=.02,
                                                           confidence=.05,
                                                           minlen=2))

inspect(or_cause_comments_rules)
```


```{r}
or_specific_cause_rules = arules::apriori(or_specific_cause_transactions, 
                                          parameter = list(support=.03, 
                                                           confidence=.05,
                                                           minlen=2))

inspect(or_specific_cause_rules)
```


```{r}
news_headlines_rules = arules::apriori(news_headlines_transactions, 
                                       parameter = list(support=.025, 
                                                        confidence=.05,
                                                        minlen=2))

inspect(news_headlines_rules)
```


## Plotting the Items Frequency

```{r}
## Plot of which items are most frequent
png(paste0("./CreatedVisuals/arm/frequency_plot_", 
           or_weather_wildfires_comments_filename, ".png"), 
      width     = 4,
      height    = 3,
      units     = "in",
      res       = 1200,
      pointsize = 4)
itemFrequencyPlot(or_cause_comments_transactions, 
                  topN=20, 
                  type="absolute", 
                  main=paste0("Top 20 Terms Frequency: ", 
                              or_weather_wildfires_comments_filename))
dev.off()
```

```{r}
## Plot of which items are most frequent
png(paste0("./CreatedVisuals/arm/frequency_plot_", 
           or_weather_wildfires_specific_filename, ".png"), 
      width     = 4,
      height    = 3,
      units     = "in",
      res       = 1200,
      pointsize = 4)
itemFrequencyPlot(or_specific_cause_transactions, 
                  topN=20, 
                  type="absolute", 
                  main=paste0("Top 20 Terms Frequency: ", 
                              or_weather_wildfires_specific_filename))
dev.off()
```

```{r}
## Plot of which items are most frequent
png(paste0("./CreatedVisuals/arm/frequency_plot_", 
           news_headlines_filename, ".png"), 
      width     = 4,
      height    = 3,
      units     = "in",
      res       = 1200,
      pointsize = 4)
itemFrequencyPlot(news_headlines_transactions, 
                  topN=20, 
                  type="absolute", 
                  main=paste0("Top 20 Terms Frequency: ", 
                              news_headlines_filename))
dev.off()
```


## Sorting


### Oregron Causes


#### Cause Comments


```{r}
## Sort rules Support
SortedRules <- sort(or_cause_comments_rules, by="support", decreasing=TRUE)
inspect(SortedRules[1:15])
(summary(SortedRules))
```

```{r}
## Sort rules Confidence
SortedRules <- sort(or_cause_comments_rules, by="confidence", decreasing=TRUE)
inspect(SortedRules[1:15])
(summary(SortedRules))
```

```{r}
## Sort rules lift
SortedRules <- sort(or_cause_comments_rules, by="lift", decreasing=TRUE)
inspect(SortedRules[1:15])
(summary(SortedRules))
```


#### Specific Causes


```{r}
## Sort rules Support
SortedRules <- sort(or_specific_cause_rules, by="support", decreasing=TRUE)
inspect(SortedRules[1:15])
(summary(SortedRules))
```

```{r}
## Sort rules Confidence
SortedRules <- sort(or_specific_cause_rules, by="confidence", decreasing=TRUE)
inspect(SortedRules[1:15])
(summary(SortedRules))
```

```{r}
## Sort rules lift
SortedRules <- sort(or_specific_cause_rules, by="lift", decreasing=TRUE)
inspect(SortedRules[1:15])
(summary(SortedRules))
```


### News Headlines


```{r}
## Sort rules Support
SortedRules <- sort(news_headlines_rules, by="support", decreasing=TRUE)
inspect(SortedRules[1:15])
(summary(SortedRules))
```

```{r}
## Sort rules Confidence
SortedRules <- sort(news_headlines_rules, by="confidence", decreasing=TRUE)
inspect(SortedRules[1:15])
(summary(SortedRules))
```

```{r}
## Sort rules lift
SortedRules <- sort(news_headlines_rules, by="lift", decreasing=TRUE)
inspect(SortedRules[1:15])
(summary(SortedRules))
```


```{r}
png(paste0("./CreatedVisuals/arm/network_", 
           or_weather_wildfires_comments_filename, ".png"), 
      width     = 8,
      height    = 4,
      units     = "in",
      res       = 1200,
      pointsize = 4)
plot(or_cause_comments_rules, method="graph")
dev.off()
```

```{r}
png(paste0("./CreatedVisuals/arm/network_", 
           or_weather_wildfires_specific_filename, ".png"), 
      width     = 15,
      height    = 3,
      units     = "in",
      res       = 1200,
      pointsize = 4)
plot(or_specific_cause_rules, method="graph")
dev.off()
```

```{r}
png(paste0("./CreatedVisuals/arm/network_", 
           news_headlines_filename, ".png"), 
      width     = 8,
      height    = 4,
      units     = "in",
      res       = 1200,
      pointsize = 4)
plot(news_headlines_rules, method="graph")
dev.off()
```
```{r}
plot(or_cause_comments_rules, method="graph", engine = "htmlwidget")
```


```{r}
plot(or_specific_cause_rules, method="graph", engine = "htmlwidget")
```


```{r}
plot(news_headlines_rules, method="graph", engine = "htmlwidget")
```


___

# Printing out Mathematics equations for the website

.
.
.
.
.


$$
\text{Given } X \rightarrow Y \text{ then,} 
$$
.
.
.
.
. 

$$
\text{                                                                                                                                                                                           }
$$


$$
\text{Support(X,Y)} 
$$

$$
= P(X \cap Y) = \frac{frequency(X, Y)}{N}
$$

$$
\text{                                                                                                                                                                                           }
$$

$$
\text{Confidence(X,Y)} 
$$

$$
= P(Y | X) = \frac{P(X \cap Y)}{P(X)} = \frac{frequency(X,Y)}{frequency(X)}
$$


$$
\text{                                                                                                                                                                                           }
$$



$$
\text{Lift(X,Y)}
$$

$$
= \frac{P(X \cap Y)}{P(Y)P(X)} = \frac{\text{Support}(X,Y)}{\text{Support}(Y)\text{Support}(X)}
$$


$$
\text{                                                                                                                                                                                           }
$$
